---
title: "Solver Benchmarks"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_caption: true
    self_contained: yes
fontsize: 11pt
documentclass: article
bibliography: references.bib
csl: reference-style.csl
vignette: >
  %\VignetteIndexEntry{Solver benchmarks}
  %\VignetteEngine{knitr::rmarkdown_notangle}
---

```{r, include = FALSE}
w <- 7
h <- 4
is_check <- ("CheckExEnv" %in% search()) || any(c("_R_CHECK_TIMINGS_",
             "_R_CHECK_LICENSE_") %in% names(Sys.getenv()))
knitr::opts_chunk$set(fig.align = "center", eval = !is_check,
                      fig.width = w, fig.height = h)
```

```{r, include = FALSE}
devtools::load_all()
```

# Introduction

TODO.

# Methods

TODO.

To start off, we will load some packages to help download the benchmark results and visualize them.

```{r "load packages", message = FALSE}
# load packages
library(prioritizr)
library(piggyback)
library(ggplot2)
library(tidyverse)
```

Next, download the benchmark results and load them into our session.

```{r "import results", message = FALSE}
# download data to temporary folder
pb_download(
  file = c("solutions.zip", "results.rda"),
  repo = "prioritizr/benchmark", dest = tempdir(), tag = "latest")

# load benchmark results
load(file.path(tempdir(), "results.rda"))

# load benchmark solutions
unzip(file.path(tempdir(), "solutions.zip"), exdir = tempdir())
solution_paths <-
  file.path(tempdir(), "solutions", paste0(benchmark_results$id, ".tif"))
solution_raster_data <- lapply(solution_paths, raster)
```

We can now inspect the benchmark results. The `benchmark_results` table contains information for each benchmark run (e.g. run time), and the `solution_raster_data` contains the solutions generated for each benchmark run.

```{r "preview results"}
# preview results
print(benchmark_results)
```

# Results

```{r "plot average run times"}
# make plot
p <-
  ggplot(benchmark_results, aes(x = solver, y = run_time)) +
  geom_boxplot()

# render plot
print(p)
```


```{r "time for pu's 1"}
# make plot
p <-
  ggplot(benchmark_results %>%
           filter(number_of_planning_units == 1478, boundary_penalty == 0), 
         aes(x = relative_target, y = run_time, color = solver)) +
  geom_line(aes(color=solver))+
  geom_point(aes(color=solver))

# render plot
print(p)
```


```{r "time for pu's 2"}
# make plot
p <-
  ggplot(benchmark_results %>%
           filter(number_of_planning_units == 12902, boundary_penalty == 0), 
         aes(x = relative_target, y = run_time, color = solver)) +
  geom_line(aes(color=solver))+
  geom_point(aes(color=solver))

# render plot
print(p)
```
```{r "time for pu's 3"}
# make plot
p <-
  ggplot(benchmark_results %>%
           filter(number_of_planning_units == 102242, boundary_penalty == 0), 
         aes(x = relative_target, y = run_time, color = solver)) +
  geom_line(aes(color=solver))+
  geom_point(aes(color=solver))

# render plot
print(p)
```
```{r "time for pu's 4"}
# make plot
p <-
  ggplot(benchmark_results %>%
           filter(number_of_planning_units == 606180, boundary_penalty == 0), 
         aes(x = relative_target, y = run_time, color = solver)) +
  geom_line(aes(color=solver))+
  geom_point(aes(color=solver))

# render plot
print(p)
```



```{r "time for pu's 1 and boundary penalty"}
# make plot
p <-
  ggplot(benchmark_results %>%
           filter(number_of_planning_units == 1478, boundary_penalty == 0.001), 
         aes(x = relative_target, y = run_time, color = solver)) +
  geom_line(aes(color=solver))+
  geom_point(aes(color=solver))

# render plot
print(p)
```


```{r "time for pu's 2 and boundary penalty"}
# make plot
p <-
  ggplot(benchmark_results %>%
           filter(number_of_planning_units == 12902, boundary_penalty == 0.001), 
         aes(x = relative_target, y = run_time, color = solver)) +
  geom_line(aes(color=solver))+
  geom_point(aes(color=solver))

# render plot
print(p)
```


```{r "time for pu's 3 and boundary penalty"}
# make plot
p <-
  ggplot(benchmark_results %>%
           filter(number_of_planning_units == 102242, boundary_penalty == 0.001), 
         aes(x = relative_target, y = run_time, color = solver)) +
  geom_line(aes(color=solver))+
  geom_point(aes(color=solver))

# render plot
print(p)
```


```{r "time for pu's 4 and boundary penalty"}
# make plot
p <-
  ggplot(benchmark_results %>%
           filter(number_of_planning_units == 606180, boundary_penalty == 0.001), 
         aes(x = relative_target, y = run_time, color = solver)) +
  geom_line(aes(color=solver))+
  geom_point(aes(color=solver))

# render plot
print(p)
```