---
title: "Calibrating trade-offs tutorial"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_caption: true
fontsize: 11pt
documentclass: article
bibliography: references.bib
csl: reference-style.csl
vignette: >
  %\VignetteIndexEntry{Calibrating trade-offs tutorial}
  %\VignetteEngine{knitr::rmarkdown_notangle}
---

```{r "knitr-dummy-variables", include = FALSE}
# define dummy variables so that vignette passes package checks
prelim_weights <- matrix(rep(NA_real_, 100), ncol = 2)
rel_tol <- matrix(rep(NA_real_, 100), ncol = 1)
topsis_results <- data.frame(
  alt.row = seq_len(3), score = runif(3), rank = seq_len(3)
)
upper_weights_limit <- NA_real_
```

```{r "knitr-config-1", include = FALSE}
# define variables for vignette figures and code execution
h <- 3.5
w <- 3.5
is_check <-
  ("CheckExEnv" %in% search()) ||
  any(c("_R_CHECK_TIMINGS_", "_R_CHECK_LICENSE_") %in% names(Sys.getenv())) ||
  !identical(Sys.getenv("MY_UNIVERSE"), "") ||
  any(c("CI", "GITHUB_ACTIONS", "GITHUB_SHA") %in% names(Sys.getenv()))
knitr::opts_chunk$set(
  fig.align = "center",
  eval = !is_check, purl = !is_check
)
# suppress progress bars
options(cli.progress_show_after = 1e6)
```

```{r "knitr-config-2", include = FALSE}
# load developmental version of package
devtools::load_all()

# set up print method
print <- function(x, ...) {
  if (!interactive() && inherits(x, "ConservationProblem")) {
    prioritizr::knit_print.ConservationProblem(x)
  } else if (!interactive() && inherits(x, "OptimizationProblem")) {
    prioritizr::knit_print.OptimizationProblem(x)
  } else {
    base::print(x)
  }
}
```

## Introduction

Systematic conservation planning requires making trade-offs between objectives [@r4; @r36]. Since different objectives may conflict with one another -- or not align perfectly -- prioritizations need to make trade-offs between different objectives [@r37]. Although some objectives can easily be accounted for by using locked constraints or representation targets [e.g., @r40; @r39], this is not always the case [e.g., @r38]. For example, prioritizations often need to balance overall cost with the overall level spatial fragmentation among priority areas [@r41; @r42]. Additionally, prioritizations often need to balance the overall level of connectivity among priority areas against other objectives [@r43]. Since the best trade-off depends on a range of factors -- such as available budgets, species' connectivity requirements, and management capacity -- finding the best compromise can be challenging.

The _prioritizr R_ package provides multi-objective optimization approaches to help identify desirable compromises between different objectives. To achieve this, one option involves formulating a conservation planning problem (via `problem()`) with a primary objective (e.g., `add_min_set_objective()` to minimize cost) and penalties specify to supplementary objectives (e.g., `add_boundary_penalties()` to minimize spatial fragmentation). Another option involves formulating a multi-objective conservation planning problem (via `multi_problem()`) with multiple objectives, (optionally) supplementary penalties, and a particular multi-objective objective optimization approach (e.g., `add_hier_approach()` for hierarchical optimization). When formulating a problem, the nature of these trade-offs can be specified using parameters (e.g., `penalty` parameter for `add_boundary_penalties()` function, or `rel_tol` parameter for `add_hier_approach()` ). To identify a prioritization that represents a desirable compromise among multiple objectives, a calibration analysis can be performed to generate a set of candidate prioritizations based on different parameters or multi-objective optimization approaches, measure their performance according to each of the objectives, and then select a prioritization based on how well it achieves the objectives [@r41; @r42; @r43].

The aim of this tutorial is to provide guidance on calibrating trade-offs when using the _prioritizr R_ package. Here we will explore a several different approaches for generating prioritizations and finding a desirable compromise between different objectives. Specifically, we will try to generate prioritizations that strike the best balance between total cost of priority areas and spatial fragmentation (measured as the total boundary length of prioritization). Although the code presented in this vignette is directly applicable to performing a boundary length calibration analysis [similar to @r45], it can also be adapted for other penalty and objective functions (e.g., exploring trade-offs between cost and species' representation).

## Data

Let's load the packages and dataset used in this tutorial. Since this tutorial uses the _prioritizrdata R_ package along with several other _R_ packages (see below), please ensure that they are all installed. This particular dataset comprises two objects: `tas_pu` and `tas_features`. Although we will briefly describe this dataset below, please refer `?prioritizrdata::tas_data` for further details.

```{r "load-data", message = FALSE}
# load packages
library(prioritizrdata)
library(prioritizr)
library(sf)
library(terra)
library(dplyr)
library(tibble)
library(ggplot2)
library(topsis)
library(withr)
library(stringr)
library(ggrepel)

# set seed for reproducibility
set.seed(500)

# load planning unit data
tas_pu <- get_tas_pu()
print(tas_pu)

# load feature data
tas_features <- get_tas_features()
print(tas_features)
```

The `tas_pu` object contains planning units represented as spatial polygons (i.e., converted to a `sf::st_sf()` object). This object has three columns that denote the following information for each planning unit: a unique identifier (`id`), unimproved land value (`cost`), and current conservation status (`locked_in`). Specifically, the conservation status column indicates if at least half the area planning unit is covered by existing protected areas (denoted by a value of 1) or not (denoted by a value of zero).

```{r "plot-pu", fig.width = w, fig.height = h}
# plot map of planning unit costs
plot(tas_pu[, "cost"])

# plot map of planning unit statuses
plot(tas_pu[, "locked_in"])
```

The `tas_features` object describes the spatial distribution of various vegetation communities (using presence/absence data). We will use these vegetation communities as the biodiversity features for the prioritization.

```{r "plot-features", fig.width = 4.5, fig.height = 4.5}
# plot map of the first four vegetation classes
plot(tas_features[[1:4]])
```

## Preliminary processing

We will now prepare the data for subsequent analysis. This is important to help make it easier to find suitable trade-off parameters, and avoid numerical scaling issues that can result in overly long run times (see `presolve_check()` for further information). These processing steps are akin to data scaling (or normalization) procedures that are applied in statistical analysis to improve model convergence.

To begin with, we will set the cost values for all locked in planning units to zero. This is important so that the total cost of the prioritization reflects the total cost of new priority areas----not total land value including existing protected areas. In other words, we want the total cost estimate for a prioritization to reflect the cost of establishing new protected areas. **This procedure is especially important for the hierarchical approach (see below), so that its trade-off parameters reflect proportionate increases in the cost of establishing new protected areas.**

```{r "process-cost", fig.width = w, fig.height = h}
# set costs for planning units covered by existing protected areas to zero
tas_pu$cost[tas_pu$locked_in > 0.5] <- 0

# plot map of planning unit costs
plot(tas_pu[, "cost"])
```

Next, we will pre-compute and maually re-scale the boundary length data. This procedure is important because boundary length values are often very high, which can cause numerical issues that result in excessive run times (see `presolve_check()` for further details).

```{r "process-bd"}
# generate boundary length data for the planning units
tas_bd <- boundary_matrix(tas_pu)

# manually re-scale the boundary length values
tas_bd <- rescale_matrix(tas_bd)
```

After completing these procedures, our data is ready for analysis.

## Initial prioritization

We will generate an initial prioritization based on our primary objective (i.e., does not account for spatial fragmentation). Specifically, we will use the minimum set objective so that the optimization process minimizes total cost. We will add representation targets to ensure that prioritizations cover 17% of each vegetation community. Additionally, we will add constraints to ensure that planning units covered by existing protected areas are selected (i.e., locked in). Finally, we will specify that the conservation planning exercise involves binary decisions (i.e., selecting or not selecting planning units for protected area establishment).

```{r "init-problem"}
# define a problem
p0 <-
  problem(tas_pu, tas_features, cost_column = "cost") %>%
  add_min_set_objective() %>%
  add_relative_targets(0.17) %>%
  add_locked_in_constraints("locked_in") %>%
  add_binary_decisions()

# print problem
print(p0)
```

```{r "init-prioritization", results = "hide", message = FALSE}
# solve problem
s0 <- solve(p0)
```

```{r "init-print"}
# print result
print(s0)
```

```{r "int-plot", fig.width = w, fig.height = h, results = "hide"}
# create column for making a map of the prioritization
s0$map_1 <- case_when(
  s0$locked_in > 0.5 ~ "locked in",
  s0$solution_1 > 0.5 ~ "priority",
  TRUE ~ "other"
)

# plot map of prioritization
plot(
  s0[, "map_1"], pal = c("purple", "grey90", "darkgreen"),
  main = NULL, key.pos = 1
)
```

We can see that the priority areas identified by the prioritization are scattered across the study area (shown in green). Indeed, relatively few priority areas are connected to existing protected areas (shown in purple) or other priority areas (shown in green). As such, the prioritization has a high level of spatial fragmentation. If it is important to avoid such levels of spatial fragmentation, then we will need to explicitly account for spatial fragmentation in the optimization process.

## Generating candidate prioritizations

Here we will start the calibration analysis by generating a set of candidate prioritizations. In particular, we will examine multiple different approaches for generating candidate prioritizations. Some of these approaches will involve generating multiple different prioritizations based on different trade-off parameters, and other approaches will involve generating a single prioritization that tries to automatically resolve trade-offs. Since this tutorial involves navigating trade-offs between the overall cost of a prioritization and the level of spatial fragmentation associated with a prioritization (as measured by total boundary length), we will generate prioritizations using different parameters related to these objectives. **Although we'll be examining multiple approaches in this tutorial, you would normally only use one of these approaches when conducting your own analysis**

### Weighted sum approach

The weighted sum approach for multi-objective optimization involves combining separate objectives (e.g., total cost and total boundary length) into a single joint objective. To achieve this, a trade-off parameter is used to specify the relative importance of each criterion. Although older versions of the _prioritizr R_ package required building multiple problems (separately) that each had their own trade-off parameter (i.e., via the `penalty` parameter for `add_boundary_penalties()`), it is now possible to automatically generate multiple prioritizations based on multiple different trade-off parameters by explicitly formulating a multi-objective optimization problem. As such, we will build a multi-objective problem (via `multi_problem()`) and use the weighted sum approach (via `add_wtd_sum_approach()`) with `weights` values to indicate trade-offs.

The main challenge with the weighted sum approach is identifying a range of suitable `weights` (or `penalty`) values to generate candidate prioritizations. If we set a `weights` value for the boundary penalties that are too low, then the penalties will have no effect (e.g., boundary length penalties would have no effect on the prioritization). Conversely, if we set a `weights` value that is too high, then the prioritization will effectively ignore the primary objective. In such cases, the prioritization will be overly spatially clustered -- because the planning unit cost values have no effect -- and contain a single reserve. Thus we need to find a suitable range of `weights` values before we can generate a set of candidate prioritizations.

We can find a suitable range of `weights` values for the boundary penalties by generating a set of preliminary prioritizations. These preliminary prioritizations will be based on different `weights` values -- similar to the process for generating the candidate prioritizations -- but solved using customized settings that sacrifice optimality for fast run times (see below for details). This is especially important because specifying a `weights` value that is too high will cause the optimization process to take a very long time to generate a solutions (due to the numerical scaling issues mentioned previously). To find a suitable range of `weights` values, we need to identify an upper limit for the `weights` value (i.e., the highest `weights` value that result in a prioritization containing a single reserve). Let's create some preliminary `weights` to identify this upper limit. **Please note that you might need to adjust the `prelim_upper` value to find the upper limit when analyzing different datasets.**

```{r "prelim-weights"}
# define a range of different weight values for the boundary penalties
## note that we use a power scale to avoid focusing too much on high weights
prelim_lower <- -5  # change this for your own data
prelim_upper <- 1.1 # change this for your own data

# create a matrix of weights
## the first column has weights fixed at 1 for the primary objective
## the second column has varying weights for the boundary penalties
prelim_weights <- matrix(
  c(
    rep(1, 9),
    round(10^seq(prelim_lower, prelim_upper, length.out = 9), 5)
  ),
  ncol = 2
)

# assign row names for preliminary weights
rownames(prelim_weights) <-
  with_options(list(scipen = 30), paste0("weight_", prelim_weights[, 2]))

# print weight values
print(prelim_weights)
```

Next, let's use the preliminary `weights` values to generate preliminary prioritizations. As mentioned earlier, we will generate these preliminary prioritizations using customized settings to reduce runtime. Specifically, we will set a time limit of 10 minutes per run, and relax the optimality gap to 20%. Although we would not normally use such settings -- because the resulting prioritizations are not guaranteed to be near-optimal (the default gap is 10%) -- this is fine here because our goal is to tune the preliminary `weights` values. Indeed, none of these preliminary prioritizations will be considered as candidate prioritizations. **Please note that you might need to set a higher time limit, or relax the optimality gap even further (e.g., 40%) when analyzing larger datasets.**

```{r "prelim-prioritizations", results = "hide", message = FALSE}
# define a multi-objective optimization problem
p1 <-
  multi_problem(
    ## primj
    cost_obj =
      problem(tas_pu, tas_features, cost_column = "cost") %>%
      add_min_set_objective() %>%
      add_relative_targets(0.17) %>%
      add_locked_in_constraints("locked_in") %>%
      add_binary_decisions(),
    boundary_obj =
      problem(tas_pu, tas_features, cost_column = "cost") %>%
      add_min_penalties_objective() %>%
      # note that we use penalty = 1 here so that trade-offs will
      # subsequently be specified by prelim_weights
      add_boundary_penalties(penalty = 1, data = tas_bd) %>%
      add_binary_decisions()
  )

# generate preliminary prioritizations based on each weight
## note that we specify a relaxed gap and time limit for the solver
prelim_wtd_sum_results <-
  p1 %>%
  add_wtd_sum_approach(weights = prelim_weights) %>%
  add_default_solver(gap = 0.2, time_limit = 10 * 60) %>%
  solve()

# rename solution columns
names(prelim_wtd_sum_results) <- str_replace_all(
  names(prelim_wtd_sum_results),
  setNames(
    rownames(prelim_weights),
    paste0("solution_", seq_len(nrow(prelim_weights)))
  )
)

# preview results
print(prelim_wtd_sum_results)
```

After generating the preliminary prioritizations, let's create some maps to visualize them. In particular, we want to understand how different weight values influence the spatial fragmentation of the prioritizations.

```{r "plot-prelim-prioritizations", fig.width = 7, fig.height = 5.0}
# plot maps of prioritizations
plot(
  x =
    prelim_wtd_sum_results %>%
    dplyr::select(starts_with("weight_")) %>%
    mutate_if(is.numeric, function(x) {
      case_when(
        prelim_wtd_sum_results$locked_in > 0.5 ~ "locked in",
        x > 0.5 ~ "priority",
        TRUE ~ "other"
      )
    }),
  pal = c("purple", "grey90", "darkgreen")
)
```

```{r "best-prelim-guess", include = FALSE}
# set best guess of upper bound for weight
upper_weight_limit <- prelim_weights[8, 2]
```

We can see that as the `weights` value used to generate the prioritizations increases, the spatial fragmentation of the prioritizations decreases. In particular, we can see that a `weights` value of `r upper_weight_limit` results in a single reserve -- meaning this is our best guess of the upper limit. Using this `weights` value as an upper limit, we will now generate a second series of prioritizations that will be the candidate prioritizations. Critically, these candidate prioritizations will not be generated using with time limit and be generated using a more suitable gap (i.e., default gap of 10%).

```{r "display-prelim-guess", results = "asis", echo = FALSE}
cat(
  paste0(
    "``` r\n",
    "# define best guess for upper weight limit\n",
    "upper_weight_limit <- ",
    upper_weight_limit,
    "\n```"
  )
)
```

```{r "wtd-sum-prioritizations", fig.width = 7, fig.height = 7.5, results = "hide", message = FALSE}
# define a new set of weights values
weights <- matrix(
  c(
    rep(1, 9),
    round(10^seq(prelim_lower, upper_weight_limit, length.out = 9), 5)
  ),
  ncol = 2
)

# assign row names for weights
rownames(weights) <-
  with_options(list(scipen = 30), paste0("weight_", weights[, 2]))

# generate prioritizations based on weight sum approach
wtd_sum_results <-
  p1 %>%
  add_wtd_sum_approach(weights = weights) %>%
  solve()

# rename columns with prioritizations so they have weight values
names(wtd_sum_results) <- str_replace_all(
  names(wtd_sum_results),
  setNames(
    rownames(weights),
    paste0("solution_", seq_len(nrow(weights)))
  )
)

# plot maps of prioritizations
plot(
  x =
    wtd_sum_results %>%
    dplyr::select(starts_with("weight_")) %>%
    mutate_if(is.numeric, function(x) {
      case_when(
        wtd_sum_results$locked_in > 0.5 ~ "locked in",
        x > 0.5 ~ "priority",
        TRUE ~ "other"
      )
    }),
  pal = c("purple", "grey90", "darkgreen")
)
```

We now have a set of candidate prioritizations generated using the weighted sum approach. The main advantages of this approach is that it is similar calibration analyses used by other decision support tools for conservation and it is relatively straightforward to implement [@r45]. However, this approach also has a key disadvantage. Because the `weights` parameter is a unitless trade-off parameter -- meaning that we can't leverage existing knowledge to specify a suitable range of `weights` values -- we first have to conduct a preliminary analysis to identify a suitable upper limit. Although finding an upper limit was fairly simple for the example dataset, it can be difficult to find for more realistic datasets with more planning units and features. In the next section, we will show how to generate a set of candidate prioritizations using the hierarchical approach -- which does not have this disadvantage.

### Hierarchical approach

The hierarchical approach for multi-objective optimization involves generating a series of incremental prioritizations -- using a different objective at each increment to refine the previous solution -- until the final solution achieves all of the objectives. The advantage with this approach is that we can specify trade-off parameters for each objective based on a percentage from optimality. This means that we can leverage our own knowledge -- or that of decision maker -- to generate a range of suitable trade-off parameters. As such, this approach -- unlike the weighted sum approach -- does not require us to generate a series of preliminary prioritizations.

This approach uses a relative tolerance (`rel_tol`) parameter to specify trade-offs. The `rel_tol` parameter specifies the relative degree -- expressed as a proportion -- to which we are willing to sacrifice higher priority objectives to better optimize lower priority objectives. As mentioned earlier, the total cost of the prioritization is the primary objective and spatial fragmentation is a supplemental objective---thus the total cost of the prioritization has a higher priority than spatial fragmentation. For example, a value of 0.05 means that we would be willing to sacrifice a 5% reduction in quality for the highest priority objective to better optimize low priority objectives. Since these values are expressed as proportions -- and not unitless values as per the weighted sum approach -- we can use domain knowledge to specify a suitable range of `rel_tol` values. For this tutorial, let's assume that it would be impractical -- per our domain knowledge -- to expend more than four times the total cost of the initial prioritization to reduce spatial fragmentation.

```{r "hier-prioritizations", fig.width = 7, fig.height = 7.5, results = "hide", message = FALSE}
# define relative tolerance values
rel_tol <- matrix(seq(0, 4, length.out = 9))

# assign row names for relative tolerance values
rownames(rel_tol) <-
  with_options(list(scipen = 30), paste0("rel_tol_", rel_tol[, 1]))

# print relative tolerance values
print(rel_tol)

# generate prioritizations based on hierarchical approach
hierarchical_results <-
  p1 %>%
  # by default, objectives are assumed to be in order of priority
  # (but this can be altered with the priority parameter)
  add_hier_approach(rel_tol = rel_tol) %>%
  solve()

# rename columns with prioritizations so they have rel_tol values
names(hierarchical_results) <- str_replace_all(
  names(hierarchical_results),
  setNames(
    rownames(rel_tol),
    paste0("solution_", seq_len(nrow(rel_tol)))
  )
)

# plot maps of prioritizations
plot(
  x =
    hierarchical_results %>%
    dplyr::select(starts_with("rel_tol_")) %>%
    mutate_if(is.numeric, function(x) {
      case_when(
        hierarchical_results$locked_in > 0.5 ~ "locked in",
        x > 0.5 ~ "priority",
        TRUE ~ "other"
      )
    }),
  pal = c("purple", "grey90", "darkgreen")
)
```

### Cohon *et al.* (1979) approach

The Cohon *et al.* [-@r49] approach aims to automatically balance trade-offs between two objectives [@r50]. Specifically, it involves generating two optimal prioritizations -- with each prioritization representing the optimal prioritization according to each criteria (e.g., total cost versus total boundary length) -- and then using performance metrics for these prioritizations to automatically derive a trade-off parameter value [@r45; @r49]. Thus, unlike the two previous approaches, this approach can be used to automatically generate a single prioritization that balances trade-offs. As such, this approach can potentially be used to find a prioritization that represents a desirable compromise in a much shorter period of time than the previous approach.

```{r "cohon-calibration", results = "hide", message = FALSE}
# create problem with boundary penalties
## note that penalty = 1 is used as a place-holder
p2 <-
  problem(tas_pu, tas_features, cost_column = "cost") %>%
  add_min_set_objective() %>%
  add_boundary_penalties(penalty = 1, data = tas_bd) %>%
  add_relative_targets(0.17) %>%
  add_locked_in_constraints("locked_in") %>%
  add_binary_decisions()

# find calibrated boundary penalty using Cohon's approach
cohon_penalty <- calibrate_cohon_penalty(p2, verbose = FALSE)
```

```{r}
# print penalty value
print(cohon_penalty[[1]])
```

Now that we have calculated a `penalty` value using this approach, we can use it to generate a prioritization.

```{r "cohon-prioritization", fig.width = w, fig.height = h, results = "hide", message = FALSE}
# generate prioritization using penalty value calculated using Cohon's approach
p3 <-
  problem(tas_pu, tas_features, cost_column = "cost") %>%
  add_min_set_objective() %>%
  add_boundary_penalties(penalty = cohon_penalty, data = tas_bd) %>%
  add_relative_targets(0.17) %>%
  add_locked_in_constraints("locked_in") %>%
  add_binary_decisions()

# solve problem
s3 <- solve(p3)

# plot map of prioritization
plot(
  x =
    s3 %>%
    mutate(
      value = case_when(
        locked_in > 0.5 ~ "locked in",
        solution_1 > 0.5 ~ "priority",
        TRUE ~ "other"
      )
    ) %>%
    dplyr::select(value),
  pal = c("purple", "grey90", "darkgreen"),
  main = NULL, key.pos = 1
)
```

### Reference point approach

The reference point approach [@r68] is another multi-objective optimization approach that can be used to balance trade-offs. Although it has a `weights` parameter that can be used to specify trade-offs among objectives (similar to the weighted sum approach), this approach can also be used to automatically identify a prioritization that equally balances how well each objective is achieved. This is because -- similar to the Cohon *et al.* [-@r49] approach -- the reference point approach considers the best and worst possible performance that could be achieved for each objective. Now let's use the reference point approach to generate a prioritization.

```{r "ref-point-prioritization", fig.width = w, fig.height = h, results = "hide", message = FALSE}
# generate prioritization based on reference point approach
s4 <-
  p1 %>%
  add_ref_point_approach() %>%
  solve()

# plot map of prioritization
plot(
  x =
    s4 %>%
    mutate(
      value = case_when(
        locked_in > 0.5 ~ "locked in",
        solution_1 > 0.5 ~ "priority",
        TRUE ~ "other"
      )
    ) %>%
    dplyr::select(value),
  pal = c("purple", "grey90", "darkgreen"),
  main = NULL, key.pos = 1
)
```

After completing this, let's compare the prioritizations to see if we can identify a desirable compromise between the objectives.

## Calculating performance metrics

Here we will calculate performance metrics to compare the prioritizations. To help organize all the prioritizations generated by the different approaches, we will first create a `solutions` object to store them in. Next, since our aim is to navigate trade-offs between the total cost of a prioritization and the overall level of spatial fragmentation associated with a prioritization (as measured by total boundary length), we will calculate metrics to assess these objectives.

```{r "metrics"}
# create object with all prioritizations
solutions <-
  sf::st_sf(geometry = sf::st_geometry(tas_pu)) %>%
  bind_cols(
    hierarchical_results %>%
      dplyr::select(starts_with("rel_tol")) %>%
      st_drop_geometry(),
    wtd_sum_results %>%
      dplyr::select(starts_with("weight")) %>%
      st_drop_geometry(),
    s3 %>%
      dplyr::select(solution_1) %>%
      rename(Cohon = solution_1) %>%
      st_drop_geometry(),
    s4 %>%
      dplyr::select(solution_1) %>%
      rename(`reference point` = solution_1) %>%
      st_drop_geometry()
  )

# calculate metrics for prioritizations
metric_data <-
  lapply(
    names(st_drop_geometry(solutions)), function(x) {
      data.frame(
        name = x,
        total_cost =
          eval_cost_summary(p0, solutions[, x])$cost,
        total_boundary_length =
          eval_boundary_summary(p0, solutions[, x])$boundary
      )
    }
  ) %>%
  bind_rows() %>%
  as_tibble() %>%
  mutate(
    approach = case_when(
      startsWith(name, "rel_tol") ~ "hierarchical",
      startsWith(name, "weight") ~ "weighted sum",
      startsWith(name, "Cohon") ~ "Cohon",
      startsWith(name, "reference point") ~ "reference point"
    )
  ) %>%
  mutate(
    label = case_when(
      approach == "hierarchical" ~
        gsub("rel_tol_", "rel_tol = ", name, fixed = TRUE),
      approach == "weighted sum" ~
        gsub("weight_", "weight = ", name, fixed = TRUE),
      TRUE ~ name
    )
  )

# preview metrics
print(metric_data)
```

After calculating the metrics, we can use them to visualize trade-offs among the prioritizations.

```{r "metric-plot", fig.width = 7, fig.height = 5.0}
# create plot to visualize trade-offs among prioritizations
result_plot <-
  ggplot(
    data = metric_data,
    aes(x = total_boundary_length, y = total_cost, label = label)
  ) +
  geom_point(aes(color = approach), size = 3) +
  geom_label_repel(
    seed = 500,
    nudge_x = 5.5,
    nudge_y = 5.5,
    force = 10,
    max.overlaps = Inf
  ) +
  scale_color_manual(
    name = "Approach",
    values = c(
      "hierarchical" = "#984ea3",
      "weighted sum" = "#000000",
      "Cohon" = "#377eb8",
      "reference point" = "#ff7f00"
    )
  ) +
  xlab("Total boundary length of prioritization") +
  ylab("Total cost of prioritization") +
  scale_x_continuous(expand = expansion(mult = c(0.2, 0.9))) +
  scale_y_continuous(expand = expansion(mult = c(0.3, 0.2))) +
  theme(
    legend.position = c(0.95, 0.95),
    legend.justification = c(1, 1)
  )

# render plot
print(result_plot)
```

## Selecting a prioritization

Now we need to decide which prioritization represents a desirable compromise between the objectives. Since weighted sum and hierarchical approaches involve generating a set of candidate prioritizations, we will begin by selecting a single prioritization from these prioritizations to help narrow down our set of choices. To achieve this, we will employ one qualitative method and one quantitative method. Although these methods could be applied to prioritizations generated with the weighted sum as well as the hierarchical approaches, here we will just consider those generated with the hierarchical approach for brevity.

### Visual method

One qualitative method involves plotting the relationship between the different criteria, and using the plot to visually select a candidate prioritization. This visual method is often used to help calibrate trade-offs among prioritizations generated using the _Marxan_ decision support tool [e.g., @r41; @r42]. In particular, we will apply the visual method to select a prioritization from the set of prioritizations generated by the hierarchical approach. So, let's create a plot to select a prioritization.

```{r "visual-method-plot", fig.width = 7, fig.height = 5.0}
# create plot to visualize trade-offs and show selected candidate prioritization
result_plot <-
  ggplot(
    data = metric_data %>% filter(approach == "hierarchical"),
    aes(x = total_boundary_length, y = total_cost, label = label)
  ) +
  geom_line() +
  geom_point(size = 3) +
  geom_label_repel(
    seed = 500,
    nudge_x = 5.5,
    nudge_y = 5.5,
    force = 10,
    max.overlaps = Inf
  ) +
  xlab("Total boundary length of prioritization") +
  ylab("Total cost of prioritization") +
  theme(
    legend.position = c(0.95, 0.95),
    legend.justification = c(1, 1)
  )

# render plot
print(result_plot)
```

We can see that there is a clear relationship between total cost and total boundary length. It would seem that in order to achieve a lower total boundary length -- and thus lower spatial fragmentation -- the prioritization must have a greater cost. Although we might expect the results to show a smoother curve -- in other words, only Pareto dominant solutions -- this result is expected because we generated candidate prioritizations using the default optimality gap of 10%. Typically, the visual method involves selecting a prioritization near the elbow (or knee) of the plot. So, let's select the prioritization generated using a `rel_tol` value of `r rel_tol[3, 1]`. To keep track of the prioritizations selected based on different methods, let's create a `method` column in the `metric_data` table.

```{r "visual-method-prioritization"}
# initialize method column
metric_data <-
  metric_data %>%
  mutate(method = "none") %>%
  mutate(
    method = if_else(
      approach %in% c("Cohon", "reference point"),
      approach,
      method
    )
  )

# specify prioritization selected by visual method
metric_data <-
  metric_data %>%
  mutate(
    method = if_else(
      name == rownames(rel_tol)[[3]],
      "visual",
      method
    )
  )
```

Next, let's consider a quantitative approach.

### TOPSIS method

Multiple-criteria decision analysis is a discipline that uses analytical methods to evaluate trade-offs between multiple objectives [MCDA; reviewed in @r44]. Although this discipline contains many different methods, here we will use the the Technique for Order of Preference by Similarity to Ideal Solution (TOPSIS) method [@r46]. This method requires (i) data describing the performance of each prioritization according to each objective, (ii) weights to encode the relative importance of each objective, and (iii) details on whether each objective should ideally be minimized or maximized. Let's run the analysis -- assuming that we want equal weighting for total cost and total boundary length -- to identify a candidate prioritization from those generated with the hierarchical approach.

```{r "topsis-method-results"}
# calculate TOPSIS scores
topsis_results <- topsis(
  decision =
    metric_data %>%
    filter(approach == "hierarchical") %>%
    dplyr::select(total_cost, total_boundary_length) %>%
    as.matrix(),
  weights = c(1, 1),
  impacts = c("-", "-")
)

# print results
print(topsis_results)
```

The candidate prioritization with the greatest TOPSIS score is considered to represent the best trade-off between total cost and total boundary length. So, based on this method, we would select the prioritization generated using a `rel_tol` value of `r rel_tol[which.max(topsis_results$score), 1]`. Let's update the `metric_data` with this information.

```{r "topsis-method-prioritization"}
# add column indicating prioritization selected by TOPSIS method
metric_data <-
  metric_data %>%
  mutate(
    method = if_else(
      name == rownames(rel_tol)[[which.max(topsis_results$score)]],
      "TOPSIS",
      method
    )
  )
```

### Method comparison

Let's create a plot to visualize the prioritizations selected by the different approaches and methods.

```{r "candidate-plot", fig.width = 7, fig.height = 5.0}
# create plot to visualize trade-offs and show selected prioritizations
result_plot <-
  ggplot(
    data = metric_data,
    aes(x = total_boundary_length, y = total_cost, label = label)
  ) +
  geom_point(aes(color = method), size = 3) +
  geom_label_repel(
    seed = 500,
    nudge_x = 5.5,
    nudge_y = 5.5,
    force = 10,
    max.overlaps = Inf
  ) +
  scale_color_manual(
    name = "Method",
    values = c(
      "visual" = "#984ea3",
      "none" = "#000000",
      "TOPSIS" = "#e41a1c",
      "Cohon" = "#377eb8",
      "reference point" = "#ff7f00"
    )
  ) +
  xlab("Total boundary length of prioritization") +
  ylab("Total cost of prioritization") +
  scale_x_continuous(expand = expansion(mult = c(0.2, 0.9))) +
  scale_y_continuous(expand = expansion(mult = c(0.3, 0.2))) +
  theme(
    legend.position = c(0.95, 0.95),
    legend.justification = c(1, 1)
  )

# render plot
print(result_plot)
```

We can see that the different methods selected different prioritizations. To further compare the results from the different methods, let's create some maps showing the selected prioritizations.

```{r "candidate-map", fig.width = 7.0, fig.height = 7.5}
# extract column names for creating the prioritizations
solutions$Visual <-
  solutions[[metric_data$name[metric_data$method == "visual"]]]
solutions$TOPSIS <-
  solutions[[metric_data$name[metric_data$method == "TOPSIS"]]]

# plot maps of selected prioritizations
plot(
  x =
    solutions %>%
    dplyr::select(Visual, TOPSIS, Cohon, `reference point`) %>%
    mutate_if(is.numeric, function(x) {
      case_when(
        tas_pu$locked_in > 0.5 ~ "locked in",
        x > 0.5 ~ "priority",
        TRUE ~ "other"
      )
    }),
  pal = c("purple", "grey90", "darkgreen")
)
```

How do we determine which one is best? This is difficult to say. Ideally, expert knowledge could be used to help select a prioritization, such as knowledge on available resources, species' connectivity requirements, and management feasibility. However, from a practical perspective, prioritizations generated for academic contexts might find the quantitative approaches more useful because they have greater transparency and reproducibility. Ultimately, all of these methods are designed to support decision making. This means that they are intended to assist the decision making process, not serve as a replacement.

## Conclusion

Hopefully, this vignette has provided a useful introduction for resolving trade-offs in prioritizations. Although we only explored trade-offs between total cost and spatial fragmentation in this tutorial, this analysis could be adapted to explore trade-offs between a wide range of different criteria. For instance, instead of considering total cost as the primary objective, future analyses could explore trade-offs with feature representation (using the `add_min_shortfall_objective()` function). Additionally, instead of spatial fragmentation, future analyses could explore trade-offs that directly relate to connectivity (using the `add_connectivity_penalties()` function) or specific variables of interest -- such as ecosystem intactness or inverse human footprint index [@r47; @r48] -- to inform decision making (using the `add_linear_penalties()` function). Furthermore, after identifying the best `weights` or `rel_tol` values to strike a balance between multiple criteria, you could generate a portfolio of prioritizations (e.g., via `add_gap_portfolio()`) to find multiple options for achieving a similar balance. This might be helpful when you need to generate a set of prioritizations that have comparable performance -- in terms of how well they achieve different criteria -- but select different planning units.

## References
