---
title: "Gurobi installation for use in prioritzr"
author: "Richard Schuster"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: true
    fig_caption: true
    self_contained: yes
fontsize: 11pt
documentclass: article
vignette: >
  %\VignetteIndexEntry{Prioritizr Basics}
  %\VignetteEngine{knitr::rmarkdown_notangle}
---

```{r, include = FALSE}
h = 3.5
w = 3.5
is_check <- ("CheckExEnv" %in% search()) || any(c("_R_CHECK_TIMINGS_",
             "_R_CHECK_LICENSE_") %in% names(Sys.getenv()))
knitr::opts_chunk$set(fig.align = "center", eval = !is_check)
```

```{r, include = FALSE}
devtools::load_all()
```

# Introduction


This vignette is part of the _prioritizr_ _R_ package for solving systematic conservation prioritization problems using integer linear programming (ILP) techniques. The package offers a flexible interface for creating conservation problems using a range of different objectives and constraints that can be tailored to the specific needs of the conservation planner. Conservation problems can be solved using a variety of commercial and open-source exact algorithm solvers. In contrast to the algorithms conventionally used to solve conservation problems, such as greedy heuristics or simulated annealing, the ILP algorithms used by _prioritizr_ are guaranteed to find optimal solutions. This package also has the functionality to read [Marxan](http://marxan.net/) input data and find much cheaper solutions in a much shorter period of time than Marxan ([Beyer _et al. 2016_](http://marxan.net/downloads/papers/beyer_etal_2015.pdf)).


For a walk-through of the functions available in the prioritizr package and their usage with a simulated dataset, refer to the [Quickstart Guide](quick_start.html).  

Please refer to the [Prioritizr Basics](prioritizr_basics.html) vignette for more on the concepts behind the _prioritizr_ package.


# The Gurobi solver

The currently supported solvers are listed below. Each must be installed separately from this package. The details of the solvers are intentionally abstracted away so that minimal knowledge is required to use a given solver. However, one of the following R packages must be installed before a conservation problem can be solved.

* [Gurobi:](http://gurobi.com) Install the [_gurobi_ _R_ package](http://www.gurobi.com/products/modeling-languages/r) to use this solver.
* [SYMPHONY:](https://projects.coin-or.org/SYMPHONY) Install either the [_Rsymphony_](https://CRAN.R-project.org/package=Rsymphony) or [_lpsymphony_](https://bioconductor.riken.jp/packages/3.3/bioc/html/lpsymphony.html) _R_ packages to use this solver.

Gurobi is the most powerful and fastest solver _prioritizr_ supports at this point, which is why we are providing the Gurobi Optimizer installation intstructions here. A great place to start in terms of software and package installation is the Gurobi [documentation](http://www.gurobi.com/documentation/)

## Gurobi installation on Windows


The R package file can be found in the <installdir>/R directory of your Gurobi installation. For a default installation of Gurobi 7.5.1, the command would be:
`install.packages('c:/gurobi751/win64/R/gurobi_7.5-1.zip', repos=NULL)`

Next you will need to install the `slam` package for gurobi to work:

`install.packages('slam','https://cloud.r-project.org')`

The next thing to try is if the R package works. To do so you can follow the [example](http://www.gurobi.com/documentation/7.5/quickstart_windows/r_example.html) from the Gurobi website, which is:

```{r}
library('gurobi')

model <- list()

model$A          <- matrix(c(1,2,3,1,1,0), nrow=2, ncol=3, byrow=T)
model$obj        <- c(1,1,2)
model$modelsense <- "max"
model$rhs        <- c(4,1)
model$sense      <- c('<', '>')
model$vtype      <- 'B'

params <- list(OutputFlag=0)

result <- gurobi(model, params)

print('Solution:')
print(result$objval)
print(result$x)
```

If R doesn't give you any error message you have successfully installed the Gurobi solver and can now use it with _prioritizr_. 

## Gurobi installation on Linux


cd /opt  
sudo tar xvfz gurobi6.5.0_linux64.tar.gz  

_add this to ~/.bashrc_  
  export GUROBI_HOME="/opt/gurobi650/linux64"  
  export PATH="${PATH}:${GUROBI_HOME}/bin"  
  export LD_LIBRARY_PATH="${GUROBI_HOME}/lib"  

rsession-ld-library-path=/opt/gurobi650/linux64/lib  


_need new licence for new install_  
grbgetkey 3e3e4daf-99be-475d-54b0-56bd060db65e  

sudo su --c "R -e \"install.packages(c('slam'), repos='https://cran.rstudio.com/')\""  
sudo su --c "R -e \"install.packages('/opt/gurobi650/linux64/R/gurobi_6.5-0_R_x86_64-unknown-linux-gnu.tar.gz', repos=NULL)\""  

_maybe use this_  
sudo chown -R rschust:rschust /opt/gurobi650  

_add to this file for Gurobi to work with RStudio Server_  
sudo nano /etc/rstudio/rserver.conf  

rsession-ld-library-path=/opt/gurobi650/linux64/lib  


## Gurobi installation on Mac OS X

TO DO

## Solving a prioritzr problem with Gurobi

If the Gurobi software and _R_ package installation on your system was successful you can test if the gurobi solver is working with the following addition to a _prioritzr_ problem forumlation:

[add_gurobi_solver()](https://prioritizr.github.io/prioritizr/reference/add_gurobi_solver.html)

Below is an exmple you can use to test if everything is working with both _prioritzr_ and the _gurobi_ package.

```{r, message = FALSE, results = "hide", fig.height = h, fig.width = w}
library(prioritizr)

# formulate the problem
p <- problem(sim_pu_raster, sim_features) %>%
      add_min_set_objective() %>%
      add_relative_targets(0.1) %>%
      add_gurobi_solver()

# solve the problem
s <- solve(p)

plot(s, col = c("grey90", "darkgreen"), main = "Solution",
     xlim = c(-0.1, 1.1), ylim = c(-0.1, 1.1))
```