---
title: "Multi-objective optimization tutorial"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_caption: true
fontsize: 11pt
documentclass: article
bibliography: references.bib
csl: reference-style.csl
vignette: >
  %\VignetteIndexEntry{Multi-objective optimization tutorial}
  %\VignetteEngine{knitr::rmarkdown_notangle}
---

```{r, include = FALSE}
h = 3.2
w = 5.0
is_check <-
  ("CheckExEnv" %in% search()) ||
  any(c("_R_CHECK_TIMINGS_", "_R_CHECK_LICENSE_") %in% names(Sys.getenv())) ||
  !identical(Sys.getenv("MY_UNIVERSE"), "") ||
  any(c("CI", "GITHUB_ACTIONS", "GITHUB_SHA") %in% names(Sys.getenv()))
knitr::opts_chunk$set(
  fig.height = h,
  fig.width = w,
  fig.align = "center",
  eval = !is_check, purl = !is_check,
  root.dir = normalizePath("../..")
)
```

```{r, include = FALSE}
# load developmental version of package
devtools::load_all()

# set up print method
print <- function(x, ...) {
  if (!interactive() && inherits(x, "ConservationProblem")) {
    prioritizr::knit_print.ConservationProblem(x)
  } else if (!interactive() && inherits(x, "OptimizationProblem")) {
    prioritizr::knit_print.OptimizationProblem(x)
  } else {
    base::print(x)
  }
}
```

## Introduction
3 paragraphs: 
- point of tutorial
- multi-objective general + concepts of pareto frontier
- vignette content

Use Washington data
around 300 sp -> assign realistic categories 

Explain some more stuff on Pareto front here but the jump straight to three objective examples
Add graph for pareto front here to explain concepts here and talk about trade-offs

## Usage

### Single-objective problem

```{r, results = "hide"}
# load packages
library(prioritizr)
library(terra)
library(tibble)

# set seed for reproducibility
set.seed(500)
```

```{r}
# import data
con_cost <- get_sim_pu_raster()
keystone_spp <- get_sim_features()[[1:3]]
iconic_spp <- get_sim_features()[[4:5]]

# set budget
con_budget <- terra::global(con_cost, "sum", na.rm = TRUE)[[1]] * 0.3

```


### Multi-objective problem

### Weighted-sum approach

### Relative constraint (hierarchical) approach

### Three objective functions

### Multiple objectives in multiple zones (EXTRA VIGNETTE FOR THIS)

## Conclusion

## References
