---
title: "Multi-objective optimization tutorial"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_caption: true
fontsize: 11pt
documentclass: article
bibliography: references.bib
csl: reference-style.csl
vignette: >
  %\VignetteIndexEntry{Multi-objective optimization tutorial}
  %\VignetteEngine{knitr::rmarkdown_notangle}
---

```{r, include = FALSE}
h = 3.2
w = 5.0
is_check <-
  ("CheckExEnv" %in% search()) ||
  any(c("_R_CHECK_TIMINGS_", "_R_CHECK_LICENSE_") %in% names(Sys.getenv())) ||
  !identical(Sys.getenv("MY_UNIVERSE"), "") ||
  any(c("CI", "GITHUB_ACTIONS", "GITHUB_SHA") %in% names(Sys.getenv()))
knitr::opts_chunk$set(
  fig.height = h,
  fig.width = w,
  fig.align = "center",
  eval = !is_check, purl = !is_check,
  root.dir = normalizePath("../..")
)
```

```{r, include = FALSE}
# load developmental version of package
devtools::load_all()

# set up print method
print <- function(x, ...) {
  if (!interactive() && inherits(x, "ConservationProblem")) {
    prioritizr::knit_print.ConservationProblem(x)
  } else if (!interactive() && inherits(x, "OptimizationProblem")) {
    prioritizr::knit_print.OptimizationProblem(x)
  } else {
    base::print(x)
  }
}
```

## Introduction

Decisions about how land, freshwater, and ocean areas are managed often involve multiple stakeholders [@rTODO]. Since different stakeholders often have their own objectives, conservation decision making often involves navigating trade-offs between competing objectives [@rTODO]. For example, stakeholders may have conservation objectives for safeguarding threatened species [@rTODO], preserving charismatic species [@rTODO], maximizing ecosystem service provisioning [@rTODO], or promoting connectivity [@rTODO]. Additionally, stakeholders may also have objectives that are not related to biodiversity conservation, such as meeting food demand through agricultural or fishing activities [@rTOD], or providing natural resources through mining activities [@rTODO]. Multi-objective optimization provides a way to explicitly evaluate multiple objectives and characterize trade-offs among them [@rTODO]. The aim of this tutorial is to show how multi-objective optimization techniques can be performed with the _prioritizr_ package. Here we will generate prioritizations based on multiple objectives and then show how these prioritizations can be used to characterize trade-offs between them and identify prioritizations that represent a desirable compromise.

## Concepts

- Multi-objective general + concepts of pareto frontier
- Explain some more stuff on Pareto front here but the jump straight to three objective examples
- Add graph for pareto front here to explain concepts here and talk about trade-offs

## Data

```{r, results = "hide"}
# load packages
library(prioritizr)
library(prioritizrdata)
library(terra)
library(tibble)
library(dplyr)

# set seed for reproducibility
set.seed(500)
```

```{r}
# import data
wa_pu <- get_wa_pu()
wa_species <- get_wa_species()
wa_carbon <- get_wa_carbon()
wa_attr <- get_wa_attr()
```

## Single-objective problem

```{r}
# calculate budget as 1% of total costs
budget <- terra::global(wa_pu, "sum", na.rm = TRUE)[[1]] * 0.01

# build problem to maximize representation of threatened species
p1 <-
  problem(wa_pu, wa_species) %>%
  add_min_shortfall_objective(budget = budget) %>%
  add_feature_weights(wa_attr$extinction_prob) %>%
  add_relative_targets(0.3) %>%
  add_binary_decisions()

# build problem to maximize representation of pubic interest species
p2 <-
  problem(wa_pu, wa_species) %>%
  add_min_shortfall_objective(budget = budget) %>%
  add_feature_weights(wa_attr$extinction_prob) %>%
  add_relative_targets(0.3) %>%
  add_binary_decisions()

# build problem to maximize preservation of vulnerable carbon
p3 <-
  problem(wa_pu, wa_carbon) %>%
  add_max_wtd_sum_objective(budget = budget) %>%
  add_binary_decisions()
```

```{r}
# define names for objectives
obj_names <- c(
  "threatened species",
  "public interest species",
  "vulnerable carbon"
)

# find solutions based on each objective (separately)
single_sols <- rast(lapply(list(p1, p2, p3), solve, run_checks = FALSE))
names(single_sols) <- obj_names

# visualize solutions
plot(single_sols)
```

## Multi-objective problem

```{r}
# build multi-objective problem
mp <- multi_problem(p1, p2, p3, problem_names = obj_names)
```

```{r}
# evaluate previously generated solutions according to each objective
single_optim_perf <-
  lapply(single_sols, eval_objective_summary, x = mp) %>%
  bind_rows() %>%
  mutate(solution = rep(obj_names, each = number_of_problems(mp)))
```

## Weighted-sum approach

```{r}
# create matrix with weights to generate variety of solutions
wts_matrix <- objective_weights_matrix(
  n_objectives = number_of_problems(mp),
  n_per_objective = 4
)

# preview weights matrix
head(wts_matrix)

# generate prioritizations with weighted sum approach
wtd_sum_sols <-
  mp %>%
  add_wtd_sum_approach(wts_matrix) %>%
  add_default_solver(verbose = FALSE) %>%
  solve(run_checks = FALSE)
```

## Hierarchical approach

```{r}
# create matrix with relative tolerance values to generate variety of solutions
rel_tol_matrix <- objective_rel_tol_matrix(
  n_objectives = number_of_problems(mp),
  n_per_objective = 4,
  max_rel_tol = 1
)

# preview relative tolerance matrix
head(rel_tol_matrix)

# generate prioritizations with hierachical appraoch
hier_sols <-
  mp %>%
  add_hier_approach(rel_tol_matrix) %>%
  add_default_solver(verbose = FALSE) %>%
  solve(run_checks = FALSE)
```

## Conclusion

## References
